---
title: "Prepare other reference datasets for CIPR"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---


# Import data


```{r}

suppressPackageStartupMessages(library(SingleR))
suppressPackageStartupMessages(library(SummarizedExperiment))



# dice <- DatabaseImmuneCellExpressionData()

mmrnaseq <- MouseRNAseqData()

# immgen <- ImmGenData()

hsrnaseq <- MonacoImmuneData()

blueprint <- BlueprintEncodeData()

hpca <- HumanPrimaryCellAtlasData()

hema <- NovershternHematopoieticData()

sumexprs <- list(# dice = dice,
                 mmrnaseq = mmrnaseq,
                 # immgen = immgen,
                 hsrnaseq = hsrnaseq, 
                 blueprint = blueprint, 
                 hpca = hpca, 
                 hema = hema)
```



# Prepare expression/sample data

## Automate processing with a function

```{r}


exprep <- function(sumexp_list){
    
    require(tibble)
    require(SummarizedExperiment)
    require(dplyr)
    
    
    exp_dat <- list()
    sample_dat <- list()
    
    
    for(i in names(sumexp_list)){
        
        obj <- sumexp_list[[i]]
        
        # Prep expression data
        dat <- as.data.frame(assay(obj))
        
        if(i == "dice") colnames(dat) <- make.names(colnames(dat), unique = T)
        
        dat_colnames <- colnames(dat)
        
        dat <- add_column(dat, Gene=rownames(dat), .after = 0)
        
        # exp_dat[[i]] <- dat
        
        # Prep sample data
        smpl <- as.data.frame(colData(obj))
        
        smpl <- smpl %>%
            add_column(short_name = rownames(smpl), .after = 0) %>%
            rename(reference_cell_type = label.main,
                   long_name = label.fine) %>%
            mutate(description = "N/A")
        
        
        
        sample_names <- smpl$short_name
        
        smpl$reference_cell_type <- factor(smpl$reference_cell_type)
        
        # sample_dat[[i]] <- smpl
        
        
        
        if(sum(dat_colnames != sample_names) != 0) message(paste("Unmatched name. Check" ,i)) else message("Column names of expression data and sample metadata is matched (before reordering).")
        
        message("reordering samples")
        
        
        reorder_vector <- order(as.character(smpl$reference_cell_type))
        
        
        
        smpl <- smpl[reorder_vector, ]
        
        sample_dat[[i]] <- smpl
        
        
        dat <- dplyr::select(dat, Gene, colnames(dat)[reorder_vector+1])
        
        exp_dat[[i]] <- dat
        
        
        new_dat_colnames <- colnames(dat)[2:dim(dat)[2]]
        new_sample_names <- smpl$short_name
        
        if(sum(new_dat_colnames != new_sample_names) != 0) message(paste("Unmatched name. Check" ,i)) else message("Column names of expression data and sample metadata is matched (after reordering).")
        
    }
    
    
    sample_dat <<- sample_dat
    exp_dat <<- exp_dat
    
}



```


## Prepare data

```{r}

exprep(sumexp_list = sumexprs)

savefun <- function(datlist, savename){
    
    for(i in names(datlist)){
        
        saveRDS(datlist[[i]], file = paste0(i, "_", savename, ".rds"))
        
        
    }
    
}


savefun(exp_dat, savename = "expr")
savefun(sample_dat, savename = "samples")

```

## Prepare data as rda files

Needed for CIPR package

```{r}


exprep(sumexp_list = sumexprs)

saverda <- function(datlist, savename){
    
    for(i in names(datlist)){
        
        assign(paste0(i, "_", savename), datlist[[i]])
        
        
        save(list = paste0(i, "_", savename),
             file = paste0(i, "_", savename, ".rda"))
        
        
    }
    
}


saverda(exp_dat, savename = "expr")
saverda(sample_dat, savename = "samples")

## test 
# load("hema_samples.rda")

```


# Reference cell type breakdown

```{r}

helper <- function(datlist){
    
    for(i in names(datlist)){
        
        
        message(i)
        
        print(
            levels(datlist[[i]]$reference_cell_type)
        )
        
    }
    
}


helper(sample_dat)




```


# Manually prepare DICE data

```{r}

# Downloaded from https://dice-database.org/downloads#expression_download


suppressPackageStartupMessages(library(dplyr))


dice_expr <- read.csv("mean_tpm_merged.csv", stringsAsFactors = F)

dice_expr[, colnames(dice_expr) != "gene"] <- log(dice_expr[, colnames(dice_expr) != "gene"]+1, base = 2)

colnames(dice_expr) <- gsub("\\.\\.", ".", colnames(dice_expr))
colnames(dice_expr) <- gsub("\\.$", "", colnames(dice_expr))

library(biomaRt)

mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
genes <- dice_expr$gene

g_list <- getBM(filters= "ensembl_gene_id",
                attributes= c("ensembl_gene_id","hgnc_symbol"),
                values=genes, 
                mart= mart)

dice_expr <- left_join(dice_expr, g_list, by=c("gene"="ensembl_gene_id"))

dice_expr <- dice_expr %>%
  dplyr::select(-gene) %>%
  dplyr::select(hgnc_symbol, everything()) %>%
  dplyr::rename(gene = hgnc_symbol) %>%
  filter(!is.na(gene))

nonzero_genes <- which(rowSums(dice_expr[, 2:dim(dice_expr)[2]])!= 0)

dice_expr <- dice_expr[nonzero_genes, ]


dice_samples <- data.frame(short_name = colnames(dice_expr)[colnames(dice_expr)!="gene"],
                         reference_cell_type = c("CD4+ T cell", 
                                                 "NK cell", 
                                                 rep("CD4+ T cell", 3),
                                                 "B cell",
                                                 rep("CD4+ T cell", 4),
                                                 rep("Monocyte",2),
                                                 "CD4+ T cell",
                                                 rep("CD8+ T cell", 2)),
                         long_name = "N/A",
                         description = "N/A")


dice_samples$short_name <- as.character(dice_samples$short_name)
dice_samples$long_name <- as.character(dice_samples$long_name)
dice_samples$description <- as.character(dice_samples$description)

save(dice_expr, file="dice_expr.rda")
save(dice_samples, file="dice_samples.rda")

saveRDS(dice_expr , "dice_expr.rds")
saveRDS(dice_samples, "dice_samples.rds")

```


# Manually prepare ImmGen data


## Download GEO Series Matrix file

This file contains multi faceted pre-normalized data including:

- Feature (gene) details
- Sample origin and preparation 

```{r, eval=F}
# Get further information about the samples
gse_v1 <- getGEO("GSE15907", GSEMatrix=T, destdir = getwd())
gse_v2 <- getGEO("GSE37448", GSEMatrix=T, destdir = getwd())

saveRDS(gse_v1, "gse_v1.rds")
saveRDS(gse_v2, "gse_v2.rds")

```


```{r}

suppressPackageStartupMessages(library(GEOquery))


gse_v1 <- readRDS("gse_v1.rds")
gse_v2 <- readRDS("gse_v2.rds")

```


## Visualize pre-normalized data

- Note that v1 and v2 normalizations were done separately and the sample medians differ.

```{r}
v1_dat <- data.frame(gse_v1$GSE15907_series_matrix.txt.gz@assayData$exprs, check.names = F)
v2_dat <- data.frame(gse_v2$GSE37448_series_matrix.txt.gz@assayData$exprs, check.names=F) 

comb_dat <- cbind(v1_dat, v2_dat)


set.seed(10)
# Separate normalizations applied
boxplot(log1p(v1_dat[,floor(runif(50, 1, dim(v1_dat)[2]))]))
boxplot(log1p(v2_dat[,floor(runif(50, 1, dim(v2_dat)[2]))]))
boxplot(log1p(comb_dat[,floor(runif(50, 1, dim(comb_dat)[2]))]))

```

## Prepare sample meta data

```{r, eval=F}

suppressPackageStartupMessages(library(dplyr))

v1_meta <- gse_v1$GSE15907_series_matrix.txt.gz@phenoData@data
v2_meta <- gse_v2$GSE37448_series_matrix.txt.gz@phenoData@data


# Uniformize gender column
colnames(v1_meta) <- gsub("Sex:ch1", "gender:ch1", colnames((v1_meta)))



cols_to_extract <- c("title",
                     "geo_accession", 
                     "description", 
                     "exclusion markers:ch1", 
                     "phenotype markers:ch1",
                     "source_name_ch1",
                     "gender:ch1",
                     "age:ch1",
                     "genetic background:ch1")

v1_meta <- v1_meta %>%
  select(!!cols_to_extract) %>%
  mutate(version = "v1")


v2_meta <- v2_meta %>%
  select(!!cols_to_extract) %>%
  mutate(version = "v2")


colnames(v1_meta) == colnames(v2_meta)

comb_meta <- rbind(v1_meta, v2_meta)

comb_meta$trimmed_title <- gsub("#[0-9]+$", "", comb_meta$title)

comb_meta$ref_cell_type <- c(rep("Pre-T cell", length(1:3)),
                                   rep("Treg", length(4:5)),
                                   rep("Pre-T cell", length(6:8)),
                                   rep("T cell", length(9:26)),
                                   rep("Pre-T cell", length(27:41)),
                                   rep("Pre-B cell", length(42:50)),
                                   rep("Stem-Progenitor", length(51:52)),
                                   rep("T cell", length(53:54)),
                                   rep("B cell", length(55:60)),
                                   rep("T cell", length(61:71)),
                                   rep("Pre-B cell", length(72:73)),
                                   rep("Stem-Progenitor", length(74:74)),
                                   rep("gd-T cell", length(75:83)),
                                   rep("Pre-B cell", length(84:86)),
                                   rep("gd-T cell", length(87:101)),
                                   rep("B cell", length(102:134)),
                                   rep("DC", length(135:179)),
                                   rep("Macrophage", length(180:182)),
                                   rep("NK cell", length(183:211)),
                                   rep("NKT cell", length(212:233)),
                                   rep("Pre-B cell", length(234:249)),
                                   rep("T cell", length(250:305)),
                                   rep("Pro-B cell", length(306:306)),
                                   rep("T cell", length(307:318)),
                                   rep("gd-T cell", length(319:339)),
                                   rep("NK cell", length(340:345)),
                                   rep("DC", length(346:365)),
                                   rep("Granulocyte", length(366:369)),
                                   rep("Macrophage", length(370:387)),
                                   rep("Monocyte", length(388:404)),
                                   rep("T cell", length(405:425)),
                                   rep("B cell", length(426:430)),
                                   rep("Stromal", length(431:468)),
                                   rep("Stem-Progenitor", length(469:500)),
                                   rep("Pre-T cell", length(501:523)),
                                   rep("B cell", length(524:526)),
                                   rep("DC", length(527:528)),
                                   rep("Macrophage", length(529:539)),
                                   rep("DC", length(540:541)),
                                   rep("Macrophage", length(542:543)),
                                   rep("DC", length(544:546)),
                                   rep("Macrophage", length(547:555)),
                                   rep("DC", length(556:561)),
                                   rep("Macrophage", length(562:564)),
                                   rep("DC", length(565:569)),
                                   rep("Stromal", length(570:570)),
                                   rep("Granulocyte", length(571:582)),
                                   rep("Macrophage", length(583:594)),
                                   rep("Monocyte", length(595:600)),
                                   rep("Pre-T cell", length(601:603)),
                                   rep("Granulocyte", length(604:606)),
                                   rep("gd-T cell", length(607:624)),
                                   rep("T cell", length(625:632)),
                                   rep("Treg", length(633:638)),
                                   rep("Macrophage", length(639:641)),
                                   rep("Stromal", length(642:650)),
                                   rep("Pre-B cell", length(651:653)),
                                   rep("gd-T cell", length(654:661)),
                                   rep("Macrophage", length(662:667)),
                                   rep("Monocyte", length(668:674)),
                                   rep("B cell", length(675:677)),
                                   rep("DC", length(678:682)),
                                   rep("Stromal", length(683:686)),
                                   rep("Eosinophil", length(687:690)),
                                   rep("Stromal", length(691:695)),
                                   rep("Granulocyte", length(696:699)),
                                   rep("Stromal", length(700:702)),
                                   rep("Macrophage", length(703:705)),
                                   rep("T cell", length(706:714)),
                                   rep("Macrophage", length(715:720)),
                                   rep("T cell", length(721:742)),
                                   rep("Monocyte", length(743:745)),
                                   rep("ILC-1", length(746:753)),
                                   rep("ILC-2", length(754:755)),
                                   rep("ILC-3", length(756:763)),
                                   rep("NK cell", length(764:771)),
                                   rep("Basophil", length(772:777)),
                                   rep("Stromal", length(778:790)),
                                   rep("Mast cell", length(791:810)),
                                   rep("Stromal", length(811:819)),
                                   rep("Macrophage", length(820:825)),
                                   rep("T cell", length(826:842)))


saveRDS(comb_meta, "combined_meta.rds")

```

## Prepare feature meta data

```{r, eval=F}

feat_v1 <- featureData(gse_v1$GSE15907_series_matrix.txt.gz) 
feat_v2 <- featureData(gse_v2$GSE37448_series_matrix.txt.gz)

# Ensure feature data are the same (same platforms)
identical(feat_v1, feat_v2)

feat_metadata <- feat_v1@data


gene_name_grabber <- function(x) {
  
  tmp <- unlist(strsplit(x, " \\/\\/ "))
  tmp[2]
}

feat_metadata$gene_id <- unname(sapply(feat_metadata$gene_assignment, gene_name_grabber))

feat_metadata$ID <- as.character(feat_metadata$ID)




```



## Analyze RAW data from CEL files


### Download CEL files

```{r, eval=F}

dir.create("CEL_files")

getGEOSuppFiles("GSE15907", baseDir = "CEL_files/")
getGEOSuppFiles("GSE37448", baseDir = "CEL_files/")

```



### Create RMA-normalized data object (instead of this, do prefiltering)


```{r, eval=F}

suppressPackageStartupMessages(library(affy))

eset <- justRMA(celfile.path = "CEL_files/")

saveRDS(eset, "eset.rds")

```

### Prepare gene expression dataset

#### Average probes targeting the same gene

```{r, eval=F}

suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(dplyr))



eset <- readRDS("eset.rds")

dat <- data.table(eset@assayData$exprs, keep.rownames = T, check.names = F)

# Shorten column names
colnames(dat) <- gsub("_.+", "", colnames(dat))

anyDuplicated(colnames(dat))


feat_annot <- fread("GPL6246-18741.txt") # Downloaded from GEO (GPL6246)


gene_info_grab <- function(x, slot_to_grab){
  
  tmp <- unlist(strsplit(x, " \\/\\/ "))
  
  tmp[slot_to_grab]
  
}

feat_annot$gene_symbol <- unname(sapply(feat_annot$gene_assignment, gene_info_grab, slot_to_grab = 2))
feat_annot$gene_description <- unname(sapply(feat_annot$gene_assignment, gene_info_grab, slot_to_grab = 3))
feat_annot$ID <- as.character(feat_annot$ID)


feat_gene_df <- select(feat_annot, ID, gene_symbol)

dat_crunch <- as.data.table(left_join(dat, feat_gene_df, by=c("rn"="ID")))

dat_crunch <- dat_crunch[, -c("rn")]

dat_crunch <- dat_crunch[ , lapply(.SD, mean), by=gene_symbol]

dat_crunch <- dat_crunch[!is.na(gene_symbol), ]

setcolorder(dat_crunch, c("gene_symbol",colnames(dat_crunch)[!(colnames(dat_crunch) %in% c("gene_symbol"))]))

```

#### Average biological replicates

```{r, eval=F}

t_dat <- transpose(dat_crunch[, -c("gene_symbol")])

colnames(t_dat) <- dat_crunch$gene_symbol

t_dat$ID <- colnames(dat_crunch)[!colnames(dat_crunch) %in% c("gene_symbol")]

setcolorder(t_dat, c("ID",colnames(t_dat)[!(colnames(t_dat) %in% c("ID"))]))



comb_meta <- readRDS("combined_meta.rds")

meta_df <- select(comb_meta, geo_accession, trimmed_title)

t_dat_crunch <- as.data.table(left_join(t_dat, meta_df, by = c("ID"="geo_accession")))

t_dat_crunch <- t_dat_crunch[, -c("ID")]

t_dat_crunch <- t_dat_crunch[, lapply(.SD, mean), by=trimmed_title]

cols <- colnames(t_dat_crunch[, -c("trimmed_title")])
rows <- t_dat_crunch$trimmed_title

immgen <- transpose(t_dat_crunch[, -c("trimmed_title")])

colnames(immgen) <- rows

immgen$GeneName <- tolower(cols)

immgen <- select(immgen, GeneName, everything())


##############################################################
saveRDS(immgen, "immgen.rds")
##############################################################


```


#### Plot overall gene expression

```{r, fig.width=16, fig.height=4}

immgen <- data.frame(readRDS("immgen.rds"), check.names = F)

boxplot(immgen[,colnames(immgen) != "GeneName"])

```


#### Ensure the gene expression profiles are retained as expected

```{r, fig.height=5, fig.width=8}

suppressPackageStartupMessages(library(data.table))

immgen <- readRDS("immgen.rds")

boxplot(immgen[GeneName == "cd3e", c("T.8Eff.Sp.OT1.d10.LisOva",
                                  "Tgd.vg3+24ahi.e17.Th",
                                  "Mo.6C+II-.BM", 
                                  "MF.II+480lo.PC",
                                  "B.Fo.Sp",
                                  "DC.103+11b-.Lu",
                                  "MC.PC")],
        las=2)

boxplot(immgen[GeneName == "cd19", c("T.8Eff.Sp.OT1.d10.LisOva",
                                  "Tgd.vg3+24ahi.e17.Th",
                                  "Mo.6C+II-.BM", 
                                  "MF.II+480lo.PC",
                                  "B.Fo.Sp",
                                  "DC.103+11b-.Lu",
                                  "MC.PC")],
        las=2)

```

## Sample metadata

```{r}

suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(dplyr))

immgen <- readRDS("immgen.rds")
comb_meta <- readRDS("combined_meta.rds")

comb_meta_dedup <- comb_meta[!duplicated(comb_meta$trimmed_title), ]



popinfo <- fread("popinfo.txt")
colnames(popinfo) <- c("short_name", "long_name", "description", "v4", "v5", "v6")

popinfo$short_name <-gsub("_", "\\.", popinfo$short_name)





samples <- colnames(immgen)[colnames(immgen) != "GeneName"]

sample_meta <- data.frame(short_name = samples) 

sample_meta2 <- left_join(sample_meta, popinfo)

sample_meta3 <- inner_join(sample_meta2, comb_meta_dedup, by=c("short_name"="trimmed_title"))


sample_meta3$description <- paste("Phenotype markers:", sample_meta3$`phenotype markers:ch1`, "//",
                                  "Exclusion markers:", sample_meta3$`exclusion markers:ch1`, "//",
                                  "Tissue:", sample_meta3$source_name_ch1, "//", 
                                  "Age:", sample_meta3$`age:ch1`, "//",
                                  "Gender:", sample_meta3$`gender:ch1`, "//",
                                  "Genetic background:", sample_meta3$`genetic background:ch1`, "//",
                                  "ImmGen release:", sample_meta3$version)

# Export sample metadata and edit manually in excel

write.csv(sample_meta3, "sample_meta3.csv")

# Read file back into R after manually editing

sample_meta3 <- read.csv("sample_meta3.csv")

immgen_annot <- select(sample_meta3, short_name, long_name, ref_cell_type, description)

colnames(immgen_annot) <- gsub("ref_", "reference_", colnames(immgen_annot)) # match the CIPR column naming

immgen_annot$long_name <- gsub("ï", "i", immgen_annot$long_name)


###################################################################
saveRDS(immgen_annot, "immgen_annot.rds")
###################################################################


ann <- readRDS("immgen_annot.rds")

ann$short_name <- as.character(ann$short_name)
ann$description <- as.character(ann$description)

saveRDS(ann, "immgen_annot.rds")

immgen_samples <- ann

save(immgen_samples, file="immgen_samples.rda")


```


## Reorder samples

```{r}

# immgen <- readRDS("immgen.rds")
# immgen_annot <- readRDS("immgen_annot.rds")

# saveRDS(immgen, "immgen_original.rds")
# saveRDS(immgen_annot, "immgen_annot_original.rds")

# immgen <- readRDS("immgen_original.rds")
# immgen_annot <- readRDS("immgen_annot_original.rds")

sum(colnames(immgen[, 2:dim(immgen)[2]]) != immgen_annot$short_name)

reorder_vector <- order(as.character(immgen_annot$reference_cell_type))

immgen_annot <- immgen_annot[reorder_vector, ]

immgen <- dplyr::select(immgen, GeneName, colnames(immgen)[reorder_vector+1])


sum(colnames(immgen[, 2:dim(immgen)[2]]) != immgen_annot$short_name)


saveRDS(immgen, "immgen.rds")
saveRDS(immgen_annot, "immgen_annot.rds")

immgen_samples <- immgen_annot
immgen_expr <- immgen

save(immgen_expr, file="immgen_expr.rda")
save(immgen_samples, file="immgen_samples.rda")

```





